---
title: "AHP Car Example"
author: "Christoph Glur"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

## Introduction

This document follows the example from [Wikipedia AHP Car Example](http://en.wikipedia.org/wiki/Analytic_hierarchy_process_%E2%80%94_Car_example). Even though a few details are different, it is required to follow along the Wikipedia example to fully understand the problem.

## What is AHP

AHP is a field of decision theory. 

## Car Example

In this example, the Jones family wants to buy a new car.

### Creating the AHP criteria tree

When deciding which car to buy, the family considers the following criteria: cost, safety, style, and capacity, and their sub-criteria. Using the ahp package, the hierarchy can be constructed like so:

```{r}
library(ahp)
root <- AhpNode$new("Car")
  cost <- root$AddChild("Cost")
    purchasePrice <- cost$AddChild("Purchase Price")
    fuelCosts <- cost$AddChild("Fuel Costs")
    maintenanceCosts <- cost$AddChild("Maintenance Costs")
    resaleValue <- cost$AddChild("Resale Value")
  safety <- root$AddChild("Safety")
    curbWeight <- safety$AddChild("Curb Weight")
    safetyClass <- safety$AddChild("Safety Class")
    crashRating <- safety$AddChild("Crash Rating")
  style <- root$AddChild("Style")
  capacity <- root$AddChild("Capacity")
    cargoCapacity <- capacity$AddChild("Cargo Capacity")
    passengerCapacity <- capacity$AddChild("Passenger Capacity")

print(as.Node(root))
```



```{r}
summary(cars)
```

You can also embed plots, for example:

```{r, echo=FALSE}
plot(cars)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
