---
title: "Tree Conversion"
author: "Christoph Glur"
date: "13.5.2015"
output: html_document
---

Converting a data.tree Node to a list is really simple:

```{r}
library(data.tree)
 
data(acme)
l <- acme$ToList()
```

Alternatively, you might prefer the "traditional" generic method:

```{r eval=FALSE}


as.list(acme)
```

Conversion to JSON becomes trivial:

```{r}
library(rjson)
cat(toJSON(l))
 
```


If we prefer our children to be in an array, we can do the following:

```{r}
cat(toJSON(acme$ToList(unname = TRUE)))
```


As a side note: you can also convert a list to a Node:

```{r}

as.Node(l)
```


There are multiple options in the ToList method. Type the following to learn more:

```{r}
?ToList
?as.list.Node
```


This opens endless possibilities for integration with other packags. For example, to draw a tree with the networkD3 package:

```{r eval=FALSE}

library(networkD3)
treeNetwork(acme$ToList(unname = TRUE))


```

```{r echo=FALSE}
library(networkD3)
treeNetwork(acme$ToList(unname = TRUE), fontSize = 16, linkColour = "steelblue", nodeColour = "steelblue", opacity = 1, margin = 100)
```

Or we could save this:

```{r}
library(htmlwidgets)
w <- treeNetwork(acme$ToList(unname = TRUE), fontSize = 16, linkColour = "steelblue", nodeColour = "steelblue", opacity = 1, margin = 100)
saveWidget(w, "acme_networkD3.html")
```

